Approximating the end of nested sampling

### 2023-08-07 paper/first_draft.tex
theory section of paper

Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
\par
Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same). The state at each iteration (i.e. the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}. At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above. For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction. 
\par
Results are shown in figure \ref{fig:toy_predictions}. For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior. The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run. This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
\par
The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly. However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance. 
\begin{figure*}
\begin{center}
	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
\end{center}
\caption{Plot showing endpoint predictions for the toy likelihoods. The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties. For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run. The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
\label{fig:toy_predictions}
\end{figure*}

\subsection{Real data}
The model was tested on real nested sampling runs for quantification of cosmological parameter tensions. Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
\par
The likelihood model clearly performs better. For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs. In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
\par
Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model. For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations. 
\begin{figure*}
\begin{center}
	\includegraphics{Figures/lcdm_predictions.pdf}
\end{center}
\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model. The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains. The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
\label{fig:lcdm_predictions}
\end{figure*}

@@

We may also rewrite the distribution \cref{eq:Pyhat} without $y$ to give
\begin{equation}
    \frac{\log\mathcal{L}_\mathrm{max}^\mathrm{live}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
\end{equation}
\begin{equation}
    P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left((\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
\end{equation}

>>>>

Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
\par
Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same). The state at each iteration (i.e. the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}. At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above. For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction. 
\par
Results are shown in figure \ref{fig:toy_predictions}. For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior. The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run. This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
\par
The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly. However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance. 
%\begin{figure*}
%\begin{center}
%	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
%	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
%	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
%\end{center}
%\caption{Plot showing endpoint predictions for the toy likelihoods. The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties. For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run. The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
%\label{fig:toy_predictions}
%\end{figure*}

\subsection{Real data}
The model was tested on real nested sampling runs for quantification of cosmological parameter tensions. Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
\par
The likelihood model clearly performs better. For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs. In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
\par
Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model. For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations. 
%\begin{figure*}
%\begin{center}
%	\includegraphics{Figures/lcdm_predictions.pdf}
%\end{center}
%\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model. The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains. The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
%\label{fig:lcdm_predictions}
%\end{figure*}

@@

We may also rewrite the distribution \cref{eq:Pyhat} without $y$ to give
\begin{equation}
    \frac{\log\mathcal{L}_\mathrm{max}^\mathrm{live}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
\end{equation}
\begin{multline}
	P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\Big( (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}\\
	-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}} \Big)^{n-1} \quad [0<y_\mathrm{max}^\mathrm{live}<1],
\end{multline}
